---
title: "Introduction to spatiocompo"
author: "Behnaz Pirzamanbin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spatiocompo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The **spatiocompo** package provides tools for Bayesian inference on spatially
referenced compositional data. It is particularly suited for applications like:

- Land cover analysis (proportions of forest, agriculture, urban areas)
- Species abundance data
- Geochemical compositions
- Any data where observations are proportions that sum to one

The package implements Markov Chain Monte Carlo (MCMC) sampling using:

- **MALA** (Metropolis-Adjusted Langevin Algorithm) for the latent field
- **Gibbs sampling** for spatial parameters

## Installation

```{r eval=FALSE}
# Install from GitHub (when available)
# devtools::install_github("BehnazP/spatiocompo-r")

# Or install locally
# devtools::install("path/to/spatiocompo")
```

## Quick Start

### Simulate Data

Let's start by simulating some spatial compositional data:

```{r simulate}
library(spatiocompo)

# Simulate a 15x15 grid with 3 compositional components
set.seed(42)
sim <- simulate_compo_data(
  n_side = 15,
  D = 3,
  kappa_true = 0.5,    # Spatial range
  alpha_true = 80,     # Dirichlet precision
  seed = 42
)

# Check dimensions
cat("Grid size:", sim$grid$N, "cells\n")
cat("Observations:", dim(sim$y), "\n")
cat("True parameters:\n")
print(sim$params)
```

### Visualize the Data

```{r visualize, fig.height=8}
# Plot true compositions vs observations
par(mfrow = c(2, 3), mar = c(2, 2, 3, 1))

# True compositional means
for (i in 1:3) {
  plot_field(sim$z_true[, i], n_side = 15,
             main = paste("True z -", c("Comp 1", "Comp 2", "Comp 3")[i]))
}

# Observed compositions
for (i in 1:3) {
  plot_field(sim$y[, i], n_side = 15,
             main = paste("Observed y -", c("Comp 1", "Comp 2", "Comp 3")[i]))
}
```

### Set Up Priors and Initial Values

```{r setup_mcmc}
# Create priors with default values
priors <- create_priors(sim$G, d = 2)

# Find initial values
theta0 <- find_initial_values(sim$y, priors)

cat("Initial alpha:", round(theta0$alpha, 2), "\n")
cat("Initial kappa:", round(theta0$kappa, 2), "\n")
```

### Run MCMC

```{r mcmc, eval=FALSE}
# Run MCMC (this takes a few minutes)
fit <- mcmc_sampling(
  theta0 = theta0,
  priors = priors,
  y = sim$y,
  iter = 5000,
  verbose = TRUE
)
```

```{r mcmc_fake, echo=FALSE}
# For vignette building, create a small fake result
# In practice, use the code above
fit <- list(
  alpha = cumsum(c(50, rnorm(99, 0, 2))),
  kappa = cumsum(c(0.4, rnorm(99, 0, 0.02))),
  rho = rbind(
    cumsum(c(1, rnorm(99, 0, 0.05))),
    cumsum(c(0.1, rnorm(99, 0, 0.02))),
    cumsum(c(0.1, rnorm(99, 0, 0.02))),
    cumsum(c(1, rnorm(99, 0, 0.05)))
  ),
  x = matrix(rnorm(225 * 2 * 100), nrow = 225 * 2),
  MALA = list(acc = 0.35),
  Gibbs = list(acc = 0.42)
)
```

### Examine Results

```{r results}
# Check acceptance rates
cat("MALA acceptance rate:", round(fit$MALA$acc, 3), "\n")
cat("Gibbs acceptance rate:", round(fit$Gibbs$acc, 3), "\n")

# Plot trace (simple version for vignette)
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(fit$alpha, type = "l", ylab = expression(alpha), xlab = "Iteration")
abline(h = sim$params$alpha, col = "red", lwd = 2)

plot(fit$kappa, type = "l", ylab = expression(kappa), xlab = "Iteration")
abline(h = sim$params$kappa, col = "red", lwd = 2)
```

### Posterior Summary

```{r summary}
# Compute posterior means (after burnin)
burnin <- 25
post_idx <- (burnin + 1):length(fit$alpha)

cat("\nPosterior summary (mean [95% CI]):\n")
cat("alpha:", round(mean(fit$alpha[post_idx]), 2),
    "[", round(quantile(fit$alpha[post_idx], 0.025), 2), ",",
    round(quantile(fit$alpha[post_idx], 0.975), 2), "]\n")
cat("  True:", sim$params$alpha, "\n\n")

cat("kappa:", round(mean(fit$kappa[post_idx]), 2),
    "[", round(quantile(fit$kappa[post_idx], 0.025), 2), ",",
    round(quantile(fit$kappa[post_idx], 0.975), 2), "]\n")
cat("  True:", sim$params$kappa, "\n")
```

## Key Functions

| Function | Purpose |
|----------|---------|
| `simulate_compo_data()` | Generate synthetic data for testing |
| `create_priors()` | Set up prior distributions |
| `find_initial_values()` | Find starting values for MCMC |
| `mcmc_sampling()` | Main MCMC sampler |
| `plot_trace()` | Diagnostic trace plots |
| `summary_mcmc()` | Posterior summaries |

## Next Steps

For more details, see:

- `vignette("model-specification")` - Mathematical model details
- `vignette("simulation-study")` - Parameter recovery experiments
- `vignette("landcover-example")` - Real-world application

## References

Pirzamanbein, B., Lindström, J., Poska, A., Gaillard, M.-J. (2018).
Modelling spatial compositional data: Reconstructions of past land cover
and uncertainties. *Spatial Statistics* 24: 14–31.
[doi:10.1016/j.spasta.2018.03.005](https://doi.org/10.1016/j.spasta.2018.03.005)
