---
title: "Model Specification"
author: "Behnaz Pirzamanbin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(spatiocompo)
```

## Overview

This vignette describes the complete mathematical model underlying the **spatiocompo**
package, based on:

> Pirzamanbein, B., Lindström, J., Poska, A., Gaillard, M.-J. (2018).
> *Modelling spatial compositional data: Reconstructions of past land
> cover and uncertainties*. Spatial Statistics 24: 14–31.
> [doi:10.1016/j.spasta.2018.03.005](https://doi.org/10.1016/j.spasta.2018.03.005)

The model combines:

1. A **Dirichlet likelihood** for compositional observations
2. A **Gaussian Markov Random Field (GMRF)** prior for spatial structure
3. An **additive log-ratio (ALR)** transformation to link the two

## The Hierarchical Model

The full hierarchical model can be written as:

$$
\begin{aligned}
\mathbf{y}_i | \mathbf{z}_i, \alpha &\sim \text{Dirichlet}(\alpha \cdot \mathbf{z}_i) \\
\mathbf{z} &= \text{inv-ALR}(\mathbf{x}) \\
\mathbf{x} | \kappa, \boldsymbol{\Sigma} &\sim \mathcal{N}(\mathbf{0}, (\boldsymbol{\Sigma}^{-1} \otimes \mathbf{Q}(\kappa))^{-1}) \\
\alpha &\sim \text{Gamma}(a, b) \\
\kappa &\sim \text{Gamma}(a_\kappa, b_\kappa) \\
\boldsymbol{\Sigma} &\sim \text{Inverse-Wishart}(\nu, \mathbf{S})
\end{aligned}
$$

where $i = 1, \ldots, n$ indexes observation locations.

## The Compositional Data Model

### Observations

Let $\mathbf{y}_i = (y_{i1}, \ldots, y_{iD})$ be a compositional observation
at location $i$, where:

- $y_{ik} \in (0, 1)$ for all components $k$
- $\sum_{k=1}^D y_{ik} = 1$

Common examples include land cover proportions (forest, grassland, urban)
or species abundances.

### Dirichlet Likelihood

We model observations using the Dirichlet distribution:

$$\mathbf{y}_i | \mathbf{z}_i, \alpha \sim \text{Dirichlet}(\alpha \cdot \mathbf{z}_i)$$

where:

- $\mathbf{z}_i = (z_{i1}, \ldots, z_{iD})$ is the underlying compositional mean
- $\alpha > 0$ is the precision parameter

The Dirichlet density is:

$$p(\mathbf{y}_i | \mathbf{z}_i, \alpha) = \frac{\Gamma(\alpha)}{\prod_{k=1}^D \Gamma(\alpha z_{ik})} \prod_{k=1}^D y_{ik}^{\alpha z_{ik} - 1}$$

**Interpretation:**

- Larger $\alpha$ means observations are concentrated near the mean $\mathbf{z}_i$
- Smaller $\alpha$ means more variability in the observations
- $\mathbb{E}[y_{ik}] = z_{ik}$
- $\text{Var}(y_{ik}) = \frac{z_{ik}(1 - z_{ik})}{\alpha + 1}$

### Complete Data Likelihood

The complete log-likelihood for all observations is:

$$\log p(\mathbf{Y} | \mathbf{Z}, \alpha) = \sum_{i=1}^n \left[ \log \Gamma(\alpha) - \sum_{k=1}^D \log \Gamma(\alpha z_{ik}) + \sum_{k=1}^D (\alpha z_{ik} - 1) \log y_{ik} \right]$$

## The Latent Gaussian Field

### ALR Transformation

To impose spatial structure while respecting the compositional constraint, we use
the **additive log-ratio (ALR)** transformation. For $d = D - 1$ components:

$$x_{ik} = \log\left(\frac{z_{ik}}{z_{iD}}\right), \quad k = 1, \ldots, d$$

The inverse transformation recovers the compositions:

$$z_{ik} = \frac{\exp(x_{ik})}{1 + \sum_{j=1}^d \exp(x_{ij})}, \quad k = 1, \ldots, d$$

$$z_{iD} = \frac{1}{1 + \sum_{j=1}^d \exp(x_{ij})}$$

```{r alr_example}
# Example: ALR transformation
z <- c(0.3, 0.5, 0.2)  # Composition (D=3)
x <- alr(z)            # ALR transform (d=2)
cat("Original composition:", z, "\n")
cat("ALR coordinates:", round(x, 3), "\n")

# Inverse recovers original
z_recovered <- invalr(x)
cat("Recovered composition:", round(z_recovered, 6), "\n")
```

### Jacobian of the Transformation

The Jacobian $\partial z / \partial x$ is needed for computing gradients. For the
ALR transformation:

$$\frac{\partial z_l}{\partial x_k} = z_l \cdot (\delta_{lk} - z_k), \quad l = 1, \ldots, d$$

$$\frac{\partial z_D}{\partial x_k} = -z_D \cdot z_k$$

where $\delta_{lk}$ is the Kronecker delta.

```{r jacobian_example}
# Jacobian computation
z <- c(0.3, 0.5, 0.2)
J <- dalr(z)
cat("Jacobian matrix (2 x 3):\n")
print(round(J, 4))
```

## GMRF Prior

### Spatial Covariance Structure

The latent field $\mathbf{x} = (x_1, \ldots, x_N)$ follows a Gaussian Markov
Random Field prior with a separable covariance structure:

$$\mathbf{x} | \kappa, \boldsymbol{\Sigma} \sim \mathcal{N}(\mathbf{0}, (\boldsymbol{\Sigma}^{-1} \otimes \mathbf{Q}(\kappa))^{-1})$$

where:

- $\kappa > 0$ is the spatial range parameter
- $\boldsymbol{\Sigma}$ is a $d \times d$ covariance matrix between components
- $\mathbf{Q}(\kappa)$ is the GMRF precision matrix
- $\otimes$ denotes the Kronecker product

This separable structure allows efficient computation by exploiting the
Kronecker product properties.

### Precision Matrix Construction

The precision matrix corresponds to a discrete approximation of a Matérn
field with smoothness $\nu = 1$ on a regular grid:

$$\mathbf{Q}(\kappa) = \kappa^4 \mathbf{I} + 2\kappa^2 \mathbf{G} + \mathbf{G}^T\mathbf{G}$$

where $\mathbf{G}$ is the discrete Laplacian matrix satisfying:

$$(\mathbf{G} \mathbf{x})_i = 4x_i - \sum_{j \sim i} x_j$$

with the sum over the four nearest neighbors (using appropriate boundary conditions).

```{r precision_example}
# Create precision matrix for a 5x5 grid
G <- create_Q(c(5, 5), alpha = 1)

# Full precision with kappa = 0.5
result <- Q_rhoxQ(NULL, kappa = 0.5, G)
Q <- result$Q

cat("Precision matrix dimensions:", dim(Q), "\n")
cat("Number of non-zeros:", sum(Q != 0), "out of", prod(dim(Q)), "\n")
cat("Sparsity:", round(100 * (1 - sum(Q != 0)/prod(dim(Q))), 1), "%\n")
```

### Effect of Kappa

The parameter $\kappa$ controls the spatial range:

- **Larger $\kappa$**: Shorter correlation range, more localized patterns
- **Smaller $\kappa$**: Longer correlation range, smoother fields

The practical range (distance at which correlation drops to ~0.1) is
approximately $\sqrt{8}/\kappa$ for $\nu = 1$.

```{r kappa_effect, fig.height=4, fig.width=7}
par(mfrow = c(1, 3), mar = c(2, 2, 3, 1))

set.seed(123)
for (kappa in c(0.1, 0.5, 1.0)) {
  sim <- simulate_compo_data(n_side = 20, kappa_true = kappa,
                             alpha_true = 1000, seed = 123)
  plot_field(sim$z_true[, 1], n_side = 20,
             main = bquote(kappa == .(kappa)))
}
```

## Prior Distributions

### Dirichlet Precision (Alpha)

$$\alpha \sim \text{Gamma}(a, b)$$

Default: $a = 1, b = 0.01$ giving a weakly informative prior with mean 100.

### Spatial Range (Kappa)

$$\kappa \sim \text{Gamma}(a_\kappa, b_\kappa)$$

Default: $a_\kappa = 2, b_\kappa = 1$ giving mean 2, mode 1.

### Component Covariance (Sigma)

$$\boldsymbol{\Sigma} \sim \text{Inverse-Wishart}(\nu, \mathbf{S})$$

Default: $\nu = d + 1$, $\mathbf{S} = \mathbf{I}_d$ (weakly informative).

With $n$ observations and the GMRF prior, the conditional posterior for
$\boldsymbol{\Sigma}$ given $\mathbf{x}$ and $\kappa$ is:

$$\boldsymbol{\Sigma} | \mathbf{x}, \kappa \sim \text{Inverse-Wishart}(\nu + N, \mathbf{S} + \mathbf{x}^T \mathbf{Q}(\kappa) \mathbf{x})$$

This conjugacy enables efficient Gibbs sampling.

## MCMC Sampling Algorithm

### Overview

The package uses a hybrid MCMC scheme:

1. **MALA** (Metropolis-Adjusted Langevin Algorithm) for $(\mathbf{x}, \alpha)$
2. **Gibbs sampling** for $(\kappa, \boldsymbol{\Sigma})$

### MALA for Latent Field and Precision

The Metropolis-Adjusted Langevin Algorithm exploits gradient information for
more efficient proposals. Let $\boldsymbol{\theta} = (\mathbf{x}, \alpha)$.

**Proposal:**
$$\boldsymbol{\theta}^* = \boldsymbol{\theta}^{(t)} + \frac{\epsilon^2}{2} \mathbf{M}^{-1} \nabla \log p(\boldsymbol{\theta}^{(t)} | \cdot) + \epsilon \mathbf{M}^{-1/2} \mathbf{z}$$

where:

- $\epsilon$ is the step size (adapted automatically)
- $\mathbf{M}$ is the Fisher information matrix (metric tensor)
- $\mathbf{z} \sim \mathcal{N}(\mathbf{0}, \mathbf{I})$

**Acceptance probability:**
$$\alpha = \min\left(1, \frac{p(\boldsymbol{\theta}^* | \cdot) \cdot q(\boldsymbol{\theta}^{(t)} | \boldsymbol{\theta}^*)}{p(\boldsymbol{\theta}^{(t)} | \cdot) \cdot q(\boldsymbol{\theta}^* | \boldsymbol{\theta}^{(t)})}\right)$$

The target acceptance rate is 0.574, which is optimal for high-dimensional
MALA under certain regularity conditions.

### Gradient of the Log-Posterior

The gradient $\nabla \log p(\mathbf{x} | \cdot)$ combines the likelihood and prior:

**Dirichlet likelihood gradient:**
$$\frac{\partial \log p(\mathbf{y} | \mathbf{z}, \alpha)}{\partial x_k} = \sum_{l=1}^D \frac{\partial \log p}{\partial z_l} \cdot \frac{\partial z_l}{\partial x_k}$$

where:
$$\frac{\partial \log p}{\partial z_l} = -\alpha \psi(\alpha z_l) + \alpha \log y_l$$

and $\psi(\cdot)$ is the digamma function.

**GMRF prior gradient:**
$$\frac{\partial \log p(\mathbf{x} | \kappa, \boldsymbol{\Sigma})}{\partial \mathbf{x}} = -(\boldsymbol{\Sigma}^{-1} \otimes \mathbf{Q}(\kappa)) \mathbf{x}$$

**Complete gradient:**
$$\nabla_\mathbf{x} \log p(\mathbf{x} | \cdot) = \mathbf{A}^T \mathbf{g} - (\boldsymbol{\Sigma}^{-1} \otimes \mathbf{Q}) \mathbf{x}$$

where $\mathbf{A}$ is the location matrix and $\mathbf{g}$ contains the
likelihood gradients.

### Fisher Information Matrix

The Fisher information (metric) combines contributions from the likelihood
and prior:

$$\mathbf{M} = -\mathbf{A}^T \mathbf{H} \mathbf{A} + \boldsymbol{\Sigma}^{-1} \otimes \mathbf{Q}(\kappa)$$

where $\mathbf{H}$ is the Hessian of the Dirichlet log-likelihood with respect
to the ALR coordinates.

### Gibbs Sampling for Spatial Parameters

Given the latent field $\mathbf{x}$:

**Step 1: Sample $\kappa$** using Metropolis-Hastings with log-normal proposal:

$$\log \kappa^* = \log \kappa^{(t)} + \epsilon \cdot z, \quad z \sim \mathcal{N}(0, 1)$$

The marginal posterior (integrating out $\boldsymbol{\Sigma}$) is:

$$p(\kappa | \mathbf{x}) \propto |\mathbf{Q}(\kappa)|^{d/2} \cdot p(\kappa) \cdot |\mathbf{S} + \mathbf{x}^T \mathbf{Q}(\kappa) \mathbf{x}|^{-(N + \nu)/2}$$

**Step 2: Sample $\boldsymbol{\Sigma}$** from its conjugate inverse-Wishart posterior:

$$\boldsymbol{\Sigma} | \mathbf{x}, \kappa \sim \text{Inverse-Wishart}(\nu + N, \mathbf{S} + \mathbf{x}^T \mathbf{Q}(\kappa) \mathbf{x})$$

This is sampled using the Bartlett decomposition.

## Efficient Computation

### Sparse Matrices

All precision matrices are stored as sparse matrices (`Matrix` package),
enabling efficient computation even for large grids.

### Fast Log-Determinants

The log-determinant $\log|\mathbf{Q}(\kappa)|$ is computed efficiently using
precomputed eigenvalues of $\mathbf{G}$:

$$\log|\mathbf{Q}(\kappa)| = \sum_{i=1}^N \log(\kappa^4 + 2\kappa^2 \lambda_i + \lambda_i^2)$$

where $\lambda_i$ are eigenvalues of $\mathbf{G}$.

```{r logdet_example}
G <- create_Q(c(10, 10), alpha = 1)
lambda <- comp_lambda(G)

# Fast computation for different kappa values
kappas <- c(0.1, 0.3, 0.5, 1.0)
for (k in kappas) {
  ld <- log_det_Q(k, lambda)
  cat("kappa =", k, ": log|Q| =", round(ld, 2), "\n")
}
```

### Cholesky Factorization

Sparse Cholesky factorization with AMD (Approximate Minimum Degree) ordering
is used for:

- Solving linear systems efficiently
- Computing log-determinants
- Generating GMRF samples

## Worked Example

```{r worked_example, fig.height=6, fig.width=7}
# Simulate spatial compositional data
set.seed(42)
sim <- simulate_compo_data(
  n_side = 10,
  D = 3,
  kappa_true = 0.5,
  alpha_true = 50,
  seed = 42
)

cat("True parameters:\n")
cat("  alpha =", sim$params$alpha, "\n")
cat("  kappa =", sim$params$kappa, "\n")
cat("  rho =\n")
print(round(sim$params$rho, 3))

# Set up priors
priors <- create_priors(sim$G, d = 2)

# Initial values
theta0 <- list(
  alpha = 30,
  kappa = 0.3,
  rho = diag(2),
  x = sim$x_true  # Start near truth for demonstration
)

# Run short MCMC (for vignette; use more iterations in practice)
fit <- mcmc_sampling(theta0, priors, sim$y, iter = 500, verbose = FALSE)

# Results
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Trace plots
plot(fit$alpha, type = "l", xlab = "Iteration", ylab = expression(alpha),
     main = "Alpha trace")
abline(h = sim$params$alpha, col = "red", lwd = 2)

plot(fit$kappa, type = "l", xlab = "Iteration", ylab = expression(kappa),
     main = "Kappa trace")
abline(h = sim$params$kappa, col = "red", lwd = 2)

# Posterior densities
hist(fit$alpha[251:500], breaks = 20, prob = TRUE,
     main = "Alpha posterior", xlab = expression(alpha))
abline(v = sim$params$alpha, col = "red", lwd = 2)

hist(fit$kappa[251:500], breaks = 20, prob = TRUE,
     main = "Kappa posterior", xlab = expression(kappa))
abline(v = sim$params$kappa, col = "red", lwd = 2)
```

```{r summary_stats}
# Posterior summary (after burn-in)
burnin <- 250
post_idx <- (burnin + 1):length(fit$alpha)

cat("\nPosterior summary (mean [95% CI]):\n\n")

# Alpha
alpha_post <- fit$alpha[post_idx]
cat("alpha:", round(mean(alpha_post), 1),
    "[", round(quantile(alpha_post, 0.025), 1), ",",
    round(quantile(alpha_post, 0.975), 1), "]\n")
cat("  True:", sim$params$alpha, "\n\n")

# Kappa
kappa_post <- fit$kappa[post_idx]
cat("kappa:", round(mean(kappa_post), 3),
    "[", round(quantile(kappa_post, 0.025), 3), ",",
    round(quantile(kappa_post, 0.975), 3), "]\n")
cat("  True:", sim$params$kappa, "\n\n")

# Acceptance rates
cat("MALA acceptance rate:", round(fit$MALA$acc, 3), "\n")
cat("Gibbs acceptance rate:", round(fit$Gibbs$acc, 3), "\n")
```

## Computational Complexity

| Operation | Complexity | Notes |
|-----------|------------|-------|
| Gradient computation | $O(n \cdot D)$ | Per observation |
| Hessian computation | $O(n \cdot D^2)$ | Sparse structure |
| Cholesky of $\mathbf{M}$ | $O(N \cdot d)$ | With AMD ordering |
| MALA proposal | $O(N \cdot d)$ | Forward/back solve |
| Log-det $\mathbf{Q}(\kappa)$ | $O(N)$ | Precomputed eigenvalues |
| IW sampling | $O(d^3)$ | Bartlett decomposition |

For a typical land cover application with $N = 1000$ grid cells and $D = 5$
components, one MCMC iteration takes approximately 10-50 ms.

## References

1. Pirzamanbein, B., Lindström, J., Poska, A., Gaillard, M.-J. (2018).
   Modelling spatial compositional data: Reconstructions of past land cover
   and uncertainties. *Spatial Statistics* 24: 14–31.
   [doi:10.1016/j.spasta.2018.03.005](https://doi.org/10.1016/j.spasta.2018.03.005)

2. Rue, H. and Held, L. (2005). *Gaussian Markov Random Fields: Theory and
   Applications*. Chapman & Hall/CRC.

3. Aitchison, J. (1986). *The Statistical Analysis of Compositional Data*.
   Chapman & Hall.

4. Girolami, M. and Calderhead, B. (2011). Riemann manifold Langevin and
   Hamiltonian Monte Carlo methods. *J. R. Statist. Soc. B*, 73(2):123-214.

5. Lindgren, F., Rue, H., and Lindström, J. (2011). An explicit link between
   Gaussian fields and Gaussian Markov random fields: the stochastic partial
   differential equation approach. *J. R. Statist. Soc. B*, 73(4):423-498.
